Remote Collection of Respiratory Samples (NACFC 2023)

### 2023-04-06 Jordana Hoppe/nacfc_2023.qmd
Update nacfc_2023.qmd

```{r}
#| include: false
library(pander)
library(tidyverse)
library(caret)
library(gtsummary)
df <- read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Jordana Hoppe/Home Collection/Data_Clean/clinical.csv", na.strings = "")
surveys <- read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Jordana Hoppe/Home Collection/Data_Clean/surveys.csv", na.strings = "")
# Combined genotype
df$Mutation.1[df$Mutation.1 != "F508del"] <- "Other"
df$Mutation.2[df$Mutation.2 != "F508del"] <- "Other"
df$Genotype <- paste(df$Mutation.1, df$Mutation.2, sep = "/")
# Age
df$Age <- difftime(df$Date.of.clinic.culture.collection, df$Patient.DOB) / 365.25
# Combine race columns
df <- df %>%
  rowwise() %>%
  mutate(
    Race = paste(
      which(c_across(Race..choice.American.Indian.or.Alaska.Native.:
      Race..choice.Unknown.or.not.reported.) == "Checked"),
      collapse = "+"
    ),
    Race = factor(Race, levels = c("", "1", "5"), labels = c("Unknown", "American Indian or Alaska Native", "White"))
  )
# Shorten organism names
colnames(df) <- sub("If.culture.was.positive..select.all.organisms.isolated...choice.", "", colnames(df))
```

# Table 1: Participant Characteristics

```{r}
#| label: tbl-1
df %>%
  select(
    Age, FEV1...predicted, FVC...predicted,
    Sex, Race, Hispanic.or.Latinx, Rural.location.
  ) %>%
  tbl_summary(
    label = list(
      Age ~ "Age at clinic culture collection",
      FEV1...predicted ~ "ppFEV", FVC...predicted ~ "ppFVC",
      Hispanic.or.Latinx ~ "Hispanic or Latinx",
      Rural.location. ~ "Rural location?"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    )
  )
```

# Sensitivity and specificity

```{r}
sens_spec <- function(gold_standard, other_test, bug, type = "Home Collection") {
  t <- table(other_test, gold_standard)
  if (all(dim(t) == c(1, 1))) {
    cat("\n")
    cat("###", bug)
    cat("\n")
    cat("#### Contingency table")
    cat("\n")
    cat("All samples marked as", unique(gold_standard))
    cat("\n")
  } else {
    sen <- sensitivity(data = factor(other_test), reference = factor(gold_standard))
    spec <- specificity(data = factor(other_test), reference = factor(gold_standard))

    dimnames(t) <- lapply(dimnames(t), function(n) {
      n <- sub("Checked", "Present", n)
      n <- sub("Unchecked", "Absent", n)
    })
    dimnames(t)$other_test <- paste(type, dimnames(t)$other_test)
    dimnames(t)$gold_standard <- paste("Clinic", dimnames(t)$gold_standard)

    cat("\n")
    cat("###", bug)
    cat("\n")
    cat("#### Contingency table")
    cat("\n")
    print(knitr::kable((as.data.frame.matrix(t))))
    cat("\n")
    cat("Sensitivity of the", tolower(type), "method was", round(sen, 3), "and specificity was", round(spec, 3))
    cat("\n")
  }
}
```

## Home Collection

```{r results='asis'}
sens_spec(df$Methicillin.sensitive.Staph.aureus.,
  df$Methicillin.sensitive.Staph.aureus..1,
  type = "Home Collection", bug = "MSSA"
)
```

```{r results='asis'}
sens_spec(df$Methicillin.resistant.Staph.aureus.,
  df$Methicillin.resistant.Staph.aureus..1,
  type = "Home Collection", bug = "MRSA"
)
```

```{r results='asis'}
sens_spec(df$A..xylosoxidans., df$A..xylosoxidans..1,
  type = "Home Collection", bug = "A. xylosoxidans"
)
```

```{r results='asis'}
sens_spec(df$Achromobacter.not.xylosoxidans., df$Achromobacter.not.xylosoxidans..1,
  type = "Home Collection", bug = "Achromobacter (not xylosoxidans)"
)
```

```{r results='asis'}
sens_spec(df$Acinetobacter., df$Acinetobacter..1,
  type = "Home Collection", bug = "Acinetobacter"
)
```

```{r results='asis'}
sens_spec(df$H..influenzae., df$H..influenzae..1,
  type = "Home Collection", bug = "H. influenzae"
)
```

```{r results='asis'}
sens_spec(df$P..aeruginosa., df$P..aeruginosa..1,
  type = "Home Collection", bug = "P. aeruginosa"
)
```

```{r results='asis'}
sens_spec(df$P..aeruginosa..mucoid.., df$P..aeruginosa..mucoid...1,
  type = "Home Collection", bug = "P. aeruginosa (mucoid)"
)
```

```{r results='asis'}
sens_spec(df$Pseudomonas.species..not.aeruginosa., df$Pseudomonas.species..not.aeruginosa..1,
  type = "Home Collection", bug = "Pseudomonas (not aeruginosa)"
)
```

```{r results='asis'}
sens_spec(df$B..cepacia.complex., df$B..cepacia.complex..1,
  type = "Home Collection", bug = "B. cepacia complex"
)
```

```{r results='asis'}
sens_spec(df$Stenotrophomonas., df$Stenotrophomonas..1,
  type = "Home Collection", bug = "Stenotrophomonas"
)
```

```{r results='asis'}
sens_spec(df$Aspergillus., df$Aspergillus..1,
  type = "Home Collection", bug = "Aspergillus"
)
```

```{r results='asis'}
sens_spec(df$Penicillium.species., df$Penicillium.species..1,
  type = "Home Collection", bug = "Penicillium species"
)
```

```{r results='asis'}
sens_spec(df$Sceosporium.species., df$Sceosporium.species..1,
  type = "Home Collection", bug = "Sceosporium species"
)
```

```{r results='asis'}
sens_spec(df$M..avium.complex., df$M..avium.complex..1,
  type = "Home Collection", bug = "M. avium complex"
)
```

```{r results='asis'}
sens_spec(df$M..abscessus., df$M..abscessus..1,
  type = "Home Collection", bug = "M. abscessus"
)
```

>>>>

```{r}
#| include: false
library(pander)
library(tidyverse)
library(caret)
library(irr)
library(gtsummary)
df <- read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Jordana Hoppe/Home Collection/Data_Clean/clinical.csv", na.strings = "")
surveys <- read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Jordana Hoppe/Home Collection/Data_Clean/surveys.csv", na.strings = "")
# Combined genotype
df$Mutation.1[df$Mutation.1 != "F508del"] <- "Other"
df$Mutation.2[df$Mutation.2 != "F508del"] <- "Other"
df$Genotype <- paste(df$Mutation.1, df$Mutation.2, sep = "/")
# Age
df$Age <- difftime(df$Date.of.clinic.culture.collection, df$Patient.DOB) / 365.25
# Combine race columns
df <- df %>%
  rowwise() %>%
  mutate(
    Race = paste(
      which(c_across(Race..choice.American.Indian.or.Alaska.Native.:
      Race..choice.Unknown.or.not.reported.) == "Checked"),
      collapse = "+"
    ),
    Race = factor(Race, levels = c("1", "5"), labels = c("American Indian or Alaska Native", "White"))
  )
# Shorten organism names
bugs = colnames(df)[grep("If.culture.was.positive..select.all.organisms.isolated...choice.",colnames(df))]
bugs <- sub("If.culture.was.positive..select.all.organisms.isolated...choice.", "", bugs)
bugs = bugs[1:16]
colnames(df) <- sub("If.culture.was.positive..select.all.organisms.isolated...choice.", "", colnames(df))
```

```{r}
#| label: tbl-1
#| tbl-cap: Participant Characteristics
df %>%
  select(
    Age, FEV1...predicted, FVC...predicted,
    Sex, Race, Hispanic.or.Latinx, Rural.location.
  ) %>%
  tbl_summary(
    label = list(
      Age ~ "Age at clinic culture collection",
      FEV1...predicted ~ "ppFEV", FVC...predicted ~ "ppFVC",
      Hispanic.or.Latinx ~ "Hispanic or Latinx",
      Rural.location. ~ "Rural location?"
    ),
    statistic = list(
      all_continuous() ~ "{mean} ({sd})"
    )
  )
```

### Agreement between all 3 methods

Fleiss' kappa is interpreted analagously to Cohen's kappa, but applies to more than two tests. Tables with NaN and NAs indicate that all tests had the same values (i.e. all absent or all present).

| Cohen's Kappa | Interpretation         |
|---------------|------------------------|
| 0             | No agreement           |
| 0.10 - 0.20   | Slight agreement       |
| 0.21 - 0.40   | Fair agreement         |
| 0.41 - 0.60   | Moderate agreement     |
| 0.61 - 0.80   | Substantial agreement  |
| 0.81 - 0.99   | Near perfect agreement |
| 1             | Perfect agreement      |

```{r}
fleiss = function(bug,bug_name){
  d = df[,c(bug,paste0(bug,".1"),paste0(bug,".2"))]
  colnames(d) = c("Clinic","Delayed","Home")
  d[d=="Checked"] = "Present"
  d[d=="Unchecked"] = "Absent"
  fk = kappam.fleiss(d)
  cat("\n")
    cat("##", bug_name)
    cat("\n")
    print(ftable(d))
    cat("\n")
    cat("Overall Fleiss' kappa for the three tests was", round(fk$value, 3), paste0("(p = ", format.pval(fk$p.value,digits = 3, eps = 0.001),")"))
    cat("\n")
}
```

```{r}
fleiss(bug = "Methicillin.sensitive.Staph.aureus.",bug_name = "MSSA")
```

```{r}
fleiss(bug = "Methicillin.resistant.Staph.aureus.",
       bug_name = "MRSA")
```

```{r}
fleiss("A..xylosoxidans.", bug_name = "A. xylosoxidans")
```

```{r}
fleiss("Achromobacter.not.xylosoxidans.",
       bug_name = "Achromobacter (not xylosoxidans)")
```

```{r}
fleiss("Acinetobacter.", bug_name = "Acinetobacter")
```

```{r}
fleiss("H..influenzae.", bug_name = "H. influenzae")
```

```{r}
fleiss("P..aeruginosa.", bug_name = "P. aeruginosa")
```

```{r}
fleiss("P..aeruginosa..mucoid..",bug_name = "P. aeruginosa (mucoid)")
```

```{r}
fleiss("Pseudomonas.species..not.aeruginosa.",
       bug_name = "Pseudomonas (not aeruginosa)")
```

```{r}
fleiss("B..cepacia.complex.", bug_name = "B. cepacia complex")
```

```{r}
fleiss("Stenotrophomonas.", bug_name = "Stenotrophomonas")
```

```{r}
fleiss("Aspergillus.",bug_name = "Aspergillus")
```

```{r}
fleiss("Penicillium.species.",bug_name = "Penicillium species")
```

```{r}
fleiss("Sceosporium.species.",bug_name = "Sceosporium species")
```

```{r}
fleiss("M..avium.complex.",bug_name = "M. avium complex")
```

```{r}
fleiss("M..abscessus.", bug_name = "M. abscessus")
```

### Figure

```{r message=FALSE}
f = df %>% 
  select(Subject.ID,contains("Methicillin.sensitive"),contains("H..influenzae."))
colnames(f) = c("ID",paste("MSSA",c("Clinic","Delayed","Home")),
                paste("H. influenzae",c("Clinic","Delayed","Home")))

counts=f %>% 
  pivot_longer(`MSSA Clinic`:`H. influenzae Home`,
                       names_to = c("Bug","Test"),
                       names_pattern = "(.*) (.*)") %>%
  filter(value=="Checked") %>%
  group_by(Bug,ID) %>%
  summarise(Test=paste0(Test,collapse = "+")) %>%
  count(Test) %>%
  mutate(prop = prop.table(n))

ggplot(counts,aes(x=Bug,y=n,fill=Test,label=n))+
  geom_col(colour = "black", position = "fill") +
  geom_text(aes(label = n), position = position_fill(vjust = 0.5)) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_brewer(palette = "Pastel1")+
  xlab("Pathogen")+ylab("% of Samples")

```
